---
title: 'BitPhylogeny: A probabilistic framework for reconstructing intra-tumor phylogenies'
output:
  html_document:
    fig_caption: yes
    number_sections: yes
  pdf_document:
    number_sections: yes
    toc: yes
  word_document: default
---


# Installation

The `BitPhylogeny` packages depends on several `python` and `R` packages. The 
first step is to make sure the following the packages are installed. 
* `python`: numpy, scipy, scikit-learn: http://scikit-learn.org/stable/, rpy2: http://rpy.sourceforge.net/ .
* `R`: rPython, mcclust, e1071, igraph, gplots, riverplot.
   
Secondly, clone the BitPhylogeny repository 
```
git clone git@bitbucket.org:ke_yuan/bitphylogeny.git 
cd bitphylogeny
```

The third step is install the BitPhylogeny python package. To do this, 
navigate into the python directory and run the following 
```
cd python
sudo python setup.py install
```
Finally, install the R package
```
cd ../R
R CMD INSTALL bitphylogenyR_0.1.tar.gz 
```
# An example
## BitPhylogeny
We use an example dataset
```{r}
library('bitphylogenyR')
example_file = system.file('sample_data.csv', package='bitphylogenyR')
tmp <- read.csv( example_file )
head(tmp)
```

Note that the last column is set to be the true cluster label of each data point. 
We separate the data and its label. 
```{r}
x <- tmp[,-dim(tmp)[2]]
true_label <- tmp[,dim(tmp)[2]]
```

Run the BitPhylogeny analysis as the following 
```{r}
bitphyloR( example_file, './output', T, 20, 10, 2)
```
The program saves the results in the directory 'output'.
```{r}
dir('./output', recursive=T)
```

The clustering performance is assessed by the V-measure. In addition, the label
trace is summarised by the maximum posterior expected adjusted Rand method.
```{r}
compute_vmeasures('./output/sample_data.csv', 
                  system.file('sample_data.csv', package='bitphylogenyR'))
```
We can visualise the results in the following way
```{r}
fp <- get_path('./output/sample_data.csv', 'mcmc-traces')
vmeasure_traces <- as.matrix(load_vmeasures(fp, 'vmeasure_traces.csv'))
mpear_vmeasure <- as.matrix(load_vmeasures(fp, 'mpear_label_vmeasure.csv'))
class(vmeasure_traces) <- 'numeric'
class(mpear_vmeasure) <- 'numeric'
par(cex.lab=1.5, cex.axis=1.5)
boxplot(vmeasure_traces, outline=F, cex.main=1.3, ylim=c(0,1),
        border=c('gray60'), col='gray90')
points(c(1,2,3),mpear_vmeasure, pch=22,cex = 1.5, bg= 'black')
colors1 <- c("gray90",'black')
colors2 <- c("gray",'black')
add_legend("bottomleft", legend=c("traces", 'BitPhylogeny'),
           pch=c(22,22), inset = c(0.1,0.20), col=colors1, 
           pt.bg=colors2,
           horiz=F, bty='n', cex=1.5)
```

The resulting trees are stored in the `treescripts` directory. The file 
`tree-freq` contains the appearance frequency of each tree in the folder. 
```{r, fig.width=15, fig.height=9}
treefreq <- read.csv('./output/sample_data.csv/treescripts//tree-freq.csv')
treefreq
plot_sankey_mft('./output/sample_data.csv/treescripts//tree-freq.csv')
```
Alternatively, the .gdl files can be visualised with the [aisee3][id] software.

## Baseline methods

The baseline methods are consisted of clustering followed by minimum spanning 
tree construction. 

### Hierarchical clustering
 
We first compute the Jacard distance matrix for the sequences.   
```{r}
dis <- dist(x, 'binary')
```
Then, we use the R function `hclust` to perform 
hierarchical clustering based on the previously computed distance matrix.
```{r}
hc <- hclust(dis)
hc
```

The resulting labels of each data point is obtained with the function `cutree`. 
The function takes the output of `hclust` and the number of desired clusters `k`. 
In this case, we set `k` to be $7$.
```{r}
label <- cutree(hc, k=7)
label
```

When there is a range of cluster number hypothesis, we compute a list of possible labels.
```{r}
K <- seq(2,14,1)
hc_cand <- lapply(K, function(ii) cutree(hc, ii) )
```

Each of these hypothesis is evaluated by the Silhouette score. 
The one with the highest score is chosen as the clustering result.
```{r}
library(cluster)
hc_silhouette_res <- sapply(1:length(K), 
                              function(ii) 
                                summary( silhouette(hc_cand[[ii]] ,dis) )$avg.width )
idx <- which.max( hc_silhouette_res )
hc_label <- hc_cand[[idx]]
hc_label
```

Once the label is computed, we compute the genotype of each cluster as the following 
```{r}
clone <- sapply(unique(hc_label), function(i) which(hc_label==i) )
n <- length(clone)
hc_genotype <- matrix(0, n, dim(x)[2])
for (i in 1:n){
    idx <- clone[[i]]
    if ( length(idx)==1 ){
      hc_genotype[i,] <- as.matrix(x[idx,])
    }else{
      hc_genotype[i,] <- as.numeric( colMeans(as.matrix(x[idx,])) > 0.5 )
    }
}
hc_genotype
```

Finally, we put the above steps into a function which gives the label and genotype estimates.
```{r}
get_label_hc
```

### K-centroids clustering
The k-centroids methods uses the same distance matrix compute above. 
```{r}
kc = pam(dis, 7)
kc
```
In this case, the genotypes can be obtained as the metroids. The Silhouette score 
is used to choose the number of clusters. 
```{r}
kc_cand <- lapply(K, function(ii) pam( dis, ii) )
kc_silhouette_res <- sapply(1:length(K), function(ii) 
                            summary( silhouette(kc_cand[[ii]]$clustering,dis) )$avg.width )
idx <- which.max( kc_silhouette_res )
  
kc_label <- kc_cand[[idx]]$clustering
kc_label
kc_genotype <- x[kc_cand[[idx]]$medoids,]
kc_genotype
```

We also wrapped up a function for k-centroids clustering. 
```{r}
get_label_kc 
```

### Tree building
We construct the minimum spanning tree based the clustering results from the previous stage. 
```{r, fig.width=15, fig.height=9}
mst <- get_mst(hc_genotype)
plot_mst(hc_genotype, hc_label, flag=F)
```

# Reproduce figure 3B
```{r}
data(saved_vmeasures)

mcmc_vmeasures <- saved_vmeasures$mcmc_vmeasures
hc_vmeasures <- saved_vmeasures$hc_vmeasures
kc_vmeasures <- saved_vmeasures$kc_vmeasures
mpear_vmeasures <- saved_vmeasures$mpear_vmeasures

par(mfrow=c(1,2), oma = c(3,3,0,0) + 0.1,
    mar = c(0,0,1,0.5) + 0.1, cex.lab=1.5, cex.axis=1.5) 
boxplot(mcmc_vmeasures$big_clone, outline=F, 
        ylim=c(0.5,1) , 
        cex.main=1.3, 
        border=c('gray60'), col='gray90')
points( mpear_vmeasures$big_clone, pch=22,cex = 1.5, bg= 'black')
points( hc_vmeasures$big_clone, pch=24,cex = 1.5, bg= 'black')
points( kc_vmeasures$big_clone, pch=25,cex = 1.5, bg= 'black') 

boxplot(mcmc_vmeasures$small_clone, outline=F, 
        ylim=c(0.5,1) , 
        yaxt='n',cex.main=1.3,
        border=c('gray60'), col='gray90')
points( mpear_vmeasures$small_clone, pch=22, cex = 1.5, bg= 'black')
points( hc_vmeasures$small_clone, pch=24, cex=1.5, bg= 'black')
points( kc_vmeasures$small_clone, pch=25, cex = 1.5, bg= 'black') 
colors1 <- c("gray90",'black', 'black', "black")
colors2 <- c("gray",'black', 'black', "black")
add_legend("bottomleft", legend=c("traces", 'BitPhylogeny',
                                  'k-centroids',
                                  'hierarchical clustering'),
           pch=c(22,22,25,24), inset = c(0.1,0.13), col=colors1, 
           pt.bg=colors2,
           horiz=F, bty='n', cex=1.5)
title(xlab = "error",
      ylab = "v-measure",
      outer = TRUE, line = 2.2)
```

#Reproduce figure 3C

```{r}
big_clone_t <- c(3, 7)
big_clone_bit <- c(2, 5)
big_clone_hc <- c(4, 7)
big_clone_kc <- c(4, 7)
big_clone_bit <- rbind(big_clone_bit, 
                      c(3, 7), c(2,9 ), c(2,7) )
big_clone_hc <- rbind(big_clone_hc, 
                     c(5, 19), c(5,20), c(6,20))
big_clone_kc <- rbind(big_clone_kc, 
                     c(5, 20), c(5, 20), c(5,20))

small_clone_t <- c(4, 12)
small_clone_bit <- c(2, 8)
small_clone_hc <- c(5, 12)
small_clone_kc <- c(5, 12)

small_clone_bit <- rbind(small_clone_bit, 
                        c(3, 14),c(2,16 ), c(2,13) )
small_clone_hc <- rbind(small_clone_hc, 
                       c(7, 19),c(5,20), c(9,20))
small_clone_kc <- rbind(small_clone_kc, 
                       c(9, 20),c(7, 20), c(6,20))

par(mfrow=c(1,2), oma = c(3,3,0,0) + 0.1,
    mar = c(0,0,1,0.5) + 0.1, cex.lab=1.5, cex.axis=1.5) 
color <- c('blue', 'red', 'red', 'red')
plot(big_clone_t[2], big_clone_t[1],pch=3, ylim=c(0,10) ,xlim= c(5,22)
     ,cex=1.5)
points(big_clone_bit[,2], big_clone_bit[,1],pch=0,cex=1.5, col=color)
points(big_clone_hc[,2], big_clone_hc[,1],pch=2,cex=1.5, col=color)
points(big_clone_kc[,2], big_clone_kc[,1],pch=6,cex=1.5, col=color)

plot(small_clone_t[2], small_clone_t[1],pch=3, ylim=c(0,10) ,xlim= c(5,22),cex=1.5,yaxt='n')
points(small_clone_bit[,2], small_clone_bit[,1],pch=0,cex=1.5, col=color)
points(small_clone_hc[,2], small_clone_hc[,1],pch=2,cex=1.5, col=color)
points(small_clone_kc[,2], small_clone_kc[,1],pch=6,cex=1.5, col=color)

colors1 <- c("black",'black', 'black', "black")
colors2 <- c("black",'black', 'black', "black")
add_legend("topleft", legend=c("truth", 'BitPhylogeny',
                                  'k-centroids',
                                  'hierarchical clustering'),
           pch=c(3,0,6,2), inset = c(0.08,0.02), col=colors1, 
           pt.bg=colors2,
           horiz=F, bty='n', cex=1.5)
add_legend("topleft", legend=c("noiseless", 'noise levels: \n0.01,0.02,0.05'),
           inset = c(0.50,0.02), text.col=c('blue','red'),
           horiz=F, bty='n', cex=1.5)
title(xlab = "number of clones",
      ylab = "maximum tree depth",
      outer = TRUE, line = 2.2)
```


[id]:http://www.absint.com/aisee/index.htm
